---
title: "PA1_template"
author: "Monica Azqueta"
date: "Friday, April 17, 2015"
output: html_document
---

## Loading and preprocessing the data

```{r}
setwd("~/Documents/Scientific Data/Reproducible research/Assigment1")
file <- "~/Documents/Scientific Data/Reproducible research/Assigment1/activity.csv"
activity_data <- read.csv(file)
```

## What is the mean total number of steps taken per day?

### 1.Calculate the total number of steps taken per day 
We make use of the function aggregate to apply the functio sum() to the variables "steps" grouped by the factor variable "date"
```{r}
steps_day <- aggregate(steps ~ date, activity_data, sum)
head(steps_day)
```

###2.Make a histogram of the total number of steps taken each day 
The histogram will show the count of days (y-axis=frequency) in which a certain total number of steps were done (x-axis). The histogram is set to have 25 bins with the argument "breaks"
```{r}
hist(steps_day$steps, breaks=seq(0, max(steps_day$steps), l=25), main = "Histogram of total number of steps each day", xlab="Total number of steps in a day")
```

###3.Calculate and report the mean and median of the total number of steps taken per day
The summary() function provides descriptive statistics for our variable steps
```{r}
summary(steps_day$steps)
```

## What is the average daily activity pattern?

###1.Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
We make use of the function aggregate() to apply the function mean to the variable "steps" grouped by the factor variable "interval"
```{r}
library(ggplot2)

steps_interval <- aggregate(steps ~ interval, activity_data, mean)

ggplot(steps_interval, aes(x=interval, y=steps)) +
  geom_line(stat="identity", colour="blue") +
  guides(fill=FALSE) + 
  labs(x="Interval [min]", y="Number of steps") +  
  labs(title="Total number of steps per time interval")
```

###2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
The function which.max provides information about where in an array the maximum values is placed
```{r}
interval_with_max_steps <- steps_interval[which.max(steps_interval$steps),]
interval_with_max_steps
```
## Imputing missing values

###1.Calculate and report the total number of missing values in the dataset 
```{r} 
missing <- sum(is.na(activity_data$steps))
missing
```

###2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. 
After replicating the "activity_data" data frame, the "NA" values in the variable "steps" are replaced for the mean of that 5-minute interval
```{r}
activity_data_filled <- activity_data
activity_data_filled$steps[is.na(activity_data$steps)]<-steps_interval$step
head(activity_data_filled)
```

###3.Make a histogram of the total number of steps taken each day.  Do these values differ from the estimates from the first part of the assignment? 
###What is the impact of imputing missing data on the estimates of the total daily number of steps?
The Histogram of the filled data frame shows, in comparison to the previous one, a higher count of days for many total number of steps per day. 
```{r}
steps_day <- aggregate(steps ~ date, activity_data_filled, sum)

hist(steps_day$steps, breaks=seq(0, max(steps_day$steps), l=25), main = "Histogram of total number of steps each day", xlab="Total number of steps in a day")
```

###Calculate and report the mean and median of the total number of steps taken per day
The descriptives statistics of the new data also show a shift in the mean value to a higher value. As expected, the median stays the same
```{r}
summary(steps_day$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

###1.Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day. 
In the graph showing the mean activity per time interval, divided in both facets (weekday and weekend), it is possibel to appreciate, that the individual performs more steps during the weekdays
```{r}
library(timeDate)

activity_data_filled$weekday<-isWeekday(activity_data_filled$date, wday = 1:5)

activity_data_filled$weekday[activity_data_filled$weekday==TRUE]<-"weekday"
activity_data_filled$weekday[activity_data_filled$weekday==FALSE]<-"weekend"

activity_data_filled$weekday<-factor(activity_data_filled$weekday)

head(activity_data_filled)
table(activity_data_filled$weekday)
```

###2.Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
steps_interval <- aggregate(steps ~ interval+weekday, activity_data_filled, mean)

ggplot(steps_interval, aes(x=interval, y=steps)) +
  geom_line(stat="identity", colour="blue", ) +
  guides(fill=TRUE) + 
  facet_grid(scales="free", space="free", .~weekday) +
  labs(x="Interval [min]", y="Number of steps") +  
  labs(title="Total number of steps per time interval") 
```